---
title: "Off-Ball Model Write Up"
authors: "Reagan Gonzales & Chaely Raynor"
format: html
editor: visual
editor_options: 
  chunk_output_type: inline
---

# Introduction

Off ball runs in soccer are one of the lesser analyzed areas of soccer analytics. In our project, we aim to answer the question, "what characteristics of an off-ball run are associated with higher threat values?"

# Data

Our data set comes from SkillCorner, and contains match, event, and tracking data from the 2023 MLS regular season.

```{r}
library(tidyverse)
set.seed(1738)
# 1. Get paths to all CSV files in the folder
file_paths <- list.files(path = "~/Library/Mobile Documents/com~apple~CloudDocs/Documents/Research/skillcorner_CMSAC2026/dynamic_events", 
                         pattern = "\\.csv$", full.names = TRUE)

# 2. Read and stack them into a single data frame
#all_matches <- read_csv(file_paths, id = "source_file")

cat("Total events loaded:", nrow(all_matches), "\n")
cat("Matches loaded:", n_distinct(all_matches$match_id), "\n")
cat("Event types:\n")
print(table(all_matches$event_type))
```

```{r}
# Only look at off-ball runs
all_matches_offball <- all_matches |>
  filter(event_type == "off_ball_run")
cat("Total off-ball run events loaded:", nrow(all_matches_offball))
```

# Adjust + drop data

```{r}
# 0.7% of xthreat values are zeros
sum(all_matches_offball$xthreat == 0, na.rm = TRUE) / 
  length(na.omit(all_matches_offball$xthreat))
# or 1761 obs
sum(all_matches_offball$xthreat == 0, na.rm = TRUE)
```

```{r}
# So mutate to 0.005
all_matches_offball <- all_matches_offball |>
  mutate(xthreat = if_else(xthreat == 0, 0.005, xthreat))
```

```{r}
# check for missing xthreat values
sum(is.na(all_matches_offball$xthreat)) # 6 missing, so drop
# drop missing data
all_matches_offball <- all_matches_offball |> drop_na(xthreat)
```

# Feature Engineering

```{r}
# Dropped off ball runs with xthreat - 0 (0.7% of data)
# Condensed off ball run types from 10 types to 4
# Made penetrating offball run baseline 
# Adjust x and y scale relative to distance to attacking goal
offball_clean <- all_matches_offball |>
  mutate(
    x_to_goal = 54.5 - x_end,   # 0 = at goal line, larger = further from goal
    y_to_goal = abs(y_end),      # 0 = central, larger = wider
  
    run_type = case_when(event_subtype %in% c( "pulling_wide","overlap", 
                                                    "underlap", 
                                                    "cross_receiver") ~ "width",
                              event_subtype %in% c("coming_short", 
                                                   "dropping_off", 
                                                   "support")  ~ "depth_dropping",
                              event_subtype %in% c("run_ahead_of_the_ball", 
                                                   "behind") ~ "penetrating",
                              event_subtype == "pulling_half_space" ~ "half_space"),
         run_type = factor(run_type, 
                           levels = c("penetrating", "width", 
                                      "depth_dropping", "half_space")))
```

```{r}
# Check factor level counts - make sure no level has too few observations
offball_clean |>
  count(run_type) |>
  mutate(pct = n / sum(n) * 100) # half space 2% of runs

offball_clean |>
  count(location_to_player_in_possession_start) |>
  mutate(pct = n / sum(n) * 100)

offball_clean |>
  count(game_state) |>
  mutate(pct = n / sum(n) * 100)
```

```{r}
# Check what attacking_side values exist
offball_clean |>
  count(attacking_side)

# Check mean x_end by attacking side
# If SkillCorner has standardized, both sides should show similar x distributions
offball_clean |>
  group_by(attacking_side) |>
  summarise(
    mean_x_end = mean(x_end, na.rm = TRUE),
    mean_x_start = mean(x_start, na.rm = TRUE),
    n = n())
```

Check for missing data

```{r}
offball_clean |>
  select(xthreat, x_end, y_end, trajectory_direction,
         run_type, speed_avg, separation_gain,
         delta_to_last_defensive_line_gain,
         distance_to_player_in_possession_end,
         location_to_player_in_possession_start,
         game_state) |>
  summarise(across(everything(), ~sum(is.na(.))))

(4793 / 227672) * 100
# 2% of speed avg is NAs, so drop
offball_clean <- offball_clean |>
  drop_na(speed_avg)

# total obs for modeling
nrow(offball_clean)
```

```{r}
# Select relevant features
offball_clean <- offball_clean |>
  select(xthreat, x_to_goal, y_to_goal, trajectory_direction,
         run_type, speed_avg, separation_gain,
         delta_to_last_defensive_line_gain,
         distance_to_player_in_possession_end,
         location_to_player_in_possession_start,
         match_id,
         n_simultaneous_runs,
         game_state)
```

# EDA

```{r}
# Distribution of xthreat
offball_clean |>
  ggplot(aes(x=xthreat)) +
  geom_histogram()
```

```{r}
# mean xthreat values for each offball run type
offball_clean |>
  group_by(event_subtype) |>
  summarize(mean_xthreat_rate = mean(xthreat, na.rm = TRUE), .groups = 'drop') |>
  ggplot(aes(x= fct_reorder(event_subtype, desc(mean_xthreat_rate)), 
             y = mean_xthreat_rate)) +
  geom_col() +
  labs(x = "", y = "Mean Xthreat Rate") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

# Model Assumptions

- **Bounded Outcome:** The dependent variable must be continuous and strictly greater than 0 and strictly less than 1.

- **Beta Distribution:** The response variable follows a conditional Beta distribution, which natively handles skewness and heteroskedasticity.

- **Linearity of the Mean:** The relationship between the predictors and a transformed mean of the response (e.g., using a logit link function) is linear.

- **Appropriate Link & Homogeneity:** Deviance residuals should be homogeneous (constant variance) when plotted against the linear predictor, and link functions must be correctly specified.

# Modeling

```{r}
library(betareg)
# cross validation set up
set.seed(25)
N_FOLDS <- 5
games_folds <- offball_clean |> 
  distinct(match_id) |>
  mutate(fold = sample(rep(1:N_FOLDS, length.out = n())))

offball_clean <- offball_clean |> 
  left_join(games_folds, by = join_by(match_id))

# cross validation
ob_cv <- function(x) {
  ob_train <- offball_clean |> filter(fold != x)
  ob_test <- offball_clean |> filter(fold == x)
  
  m1_beta_fit <- betareg(xthreat ~ x_to_goal + y_to_goal,
                         data = ob_train,
                         link = "logit")
  m2_beta_fit <- betareg(xthreat ~ x_to_goal + y_to_goal + run_type + speed_avg,
                         data = ob_train,
                         link = "logit")
  m3_beta_fit <- betareg(xthreat ~ x_to_goal + y_to_goal + run_type + 
                           speed_avg + separation_gain +
                           delta_to_last_defensive_line_gain,
                         data = ob_train,
                         link = "logit")
  m4_beta_fit <- betareg(xthreat ~ x_to_goal + y_to_goal + run_type +
                           speed_avg + separation_gain + 
                           delta_to_last_defensive_line_gain +
                           distance_to_player_in_possession_end +
                           location_to_player_in_possession_start +
                           n_simultaneous_runs + game_state,
                         data = ob_train,
                         link = "logit")
  
  ob_out <- tibble(
    m1_pred = predict(m1_beta_fit, newdata = ob_test, type = "response"),
    m2_pred = predict(m2_beta_fit, newdata = ob_test, type = "response"),
    m3_pred = predict(m3_beta_fit, newdata = ob_test, type = "response"),
    m4_pred = predict(m4_beta_fit, newdata = ob_test, type = "response"),
    test_actual = ob_test$xthreat,
    test_fold = x
  )
  return(ob_out)  
}

ob_preds <- map(1:N_FOLDS, ob_cv) |> 
  list_rbind()
```

```{r}
# cv rmse + mae
ob_preds |>
  pivot_longer(
    cols = m1_pred:m4_pred,
    names_to = "model",
    values_to = "test_pred") |>
  group_by(model, test_fold) |>
  summarize(
    rmse = sqrt(mean((test_pred - test_actual)^2)),
    mae = mean(abs(test_pred - test_actual)),
    .groups = "drop") |>
  group_by(model) |>
  summarize(
    cv_rmse = mean(rmse),
    se_rmse = sd(rmse) / sqrt(N_FOLDS),
    cv_mae = mean(mae),
    se_mae = sd(mae) / sqrt(N_FOLDS))
```

Model 4 performed best (lowest RMSE and MAE)

```{r}
# CV best beta model with linear regression
# cross validation set up
set.seed(25)
N_FOLDS <- 5
games_folds <- offball_clean |> 
  distinct(match_id) |>
  mutate(fold = sample(rep(1:N_FOLDS, length.out = n())))

offball_clean <- offball_clean |> 
  left_join(games_folds, by = join_by(match_id))

# cross validation
ob_cv <- function(x) {
  ob_train <- offball_clean |> filter(fold != x)
  ob_test <- offball_clean |> filter(fold == x)
  
  m4_beta_fit <- betareg(xthreat ~ x_to_goal + y_to_goal + run_type +
                           speed_avg + separation_gain + 
                           delta_to_last_defensive_line_gain +
                           distance_to_player_in_possession_end +
                           location_to_player_in_possession_start +
                           n_simultaneous_runs + game_state,
                         data = ob_train,
                         link = "logit")
  m5_linear_fit <- lm(log(xthreat) ~ x_to_goal + y_to_goal + run_type +
                        speed_avg + separation_gain + 
                        delta_to_last_defensive_line_gain +
                        distance_to_player_in_possession_end +
                        location_to_player_in_possession_start +
                        n_simultaneous_runs + game_state,
                      data = ob_train)
  
  ob_out <- tibble(
    m4_pred = predict(m4_beta_fit, newdata = ob_test, type = "response"),
    m5_pred = predict(m5_linear_fit, newdata = ob_test, type = "response"),
    test_actual = ob_test$xthreat,
    test_fold = x
  )
  return(ob_out)  
}

ob_preds <- map(1:N_FOLDS, ob_cv) |> 
  list_rbind()
```

```{r}
# cv linear and beta
ob_preds |>
  pivot_longer(
    cols = m4_pred:m5_pred,
    names_to = "model",
    values_to = "test_pred") |>
  group_by(model, test_fold) |>
  summarize(
    rmse = sqrt(mean((test_pred - test_actual)^2)),
    mae = mean(abs(test_pred - test_actual)),
    .groups = "drop") |>
  group_by(model) |>
  summarize(
    cv_rmse = mean(rmse),
    se_rmse = sd(rmse) / sqrt(N_FOLDS),
    cv_mae = mean(mae),
    se_mae = sd(mae) / sqrt(N_FOLDS))
```

Model 4 still performs better

```{r}
# beta GAM vs beta regression
library(mgcv)
# cross validation set up
set.seed(25)
N_FOLDS <- 5
games_folds <- offball_clean |> 
  distinct(match_id) |>
  mutate(fold = sample(rep(1:N_FOLDS, length.out = n())))

offball_clean <- offball_clean |> 
  left_join(games_folds, by = join_by(match_id))

# cross validation
ob_cv <- function(x) {
  ob_train <- offball_clean |> filter(fold != x)
  ob_test <- offball_clean |> filter(fold == x)
  
  m4_beta_fit <- betareg(xthreat ~ x_to_goal + y_to_goal + run_type +
                           speed_avg + separation_gain + 
                           delta_to_last_defensive_line_gain +
                           distance_to_player_in_possession_end +
                           location_to_player_in_possession_start +
                           n_simultaneous_runs + game_state,
                         data = ob_train,
                         link = "logit")
  m5_beta_gam <- gam(xthreat ~ s(x_to_goal) + s(y_to_goal) + run_type + s(speed_avg) + 
                       s(separation_gain) + s(delta_to_last_defensive_line_gain) +
                       s(distance_to_player_in_possession_end) + 
                       location_to_player_in_possession_start + 
                       s(n_simultaneous_runs, k=5) +
                       game_state,
                     family = betar(link = "logit"),
                     method = "REML", # more stable solution than default
                     data = ob_train)
  
  ob_out <- tibble(
    m4_pred = predict(m4_beta_fit, newdata = ob_test, type = "response"),
    m5_pred = predict(m5_beta_gam, newdata = ob_test, type = "response"),
    test_actual = ob_test$xthreat,
    test_fold = x
  )
  return(ob_out)  
}

ob_preds <- map(1:N_FOLDS, ob_cv) |> 
  list_rbind()
```

```{r}
# cv beta reg and beta gam
ob_preds |>
  pivot_longer(
    cols = m4_pred:m5_pred,
    names_to = "model",
    values_to = "test_pred") |>
  group_by(model, test_fold) |>
  summarize(
    rmse = sqrt(mean((test_pred - test_actual)^2)),
    mae = mean(abs(test_pred - test_actual)),
    .groups = "drop") |>
  group_by(model) |>
  summarize(
    cv_rmse = mean(rmse),
    se_rmse = sd(rmse) / sqrt(N_FOLDS),
    cv_mae = mean(mae),
    se_mae = sd(mae) / sqrt(N_FOLDS))
```

# Interpretations

Evaluating a beta regression model in R (typically fit with the `betareg` package) requires checking **goodness-of-fit**, **predictive accuracy**, and **model assumptions** since data is bounded between 0 and 1. \[[1](https://topmodels.r-forge.r-project.org/betareg/vignettes/betareg.html), [2](https://r-statistics.co/Beta-Regression-in-R.html), [3](https://www.youtube.com/watch?v=oR5W1r4cKmA&t=17)\]

1.  **Pseudo R²:** Standard R² doesn't apply. Instead, use the R² returned by `summary(model)` which reflects the squared correlation between the linear predictor and the transformed response. \[[1](https://r-statistics.co/Beta-Regression-in-R.html)\]

2.  **Information Criteria:** Compare models using `AIC(model)` or `BIC(model)`. Lower values indicate a better balance of fit and complexity.

3.  **Residual Diagnostics:** Use the `plot(model)` function to check assumptions. Specifically, examine the **worm plot** or **Q-Q plot** of quantile residuals to ensure errors are normally distributedn

# Discussion/conclusion

# Limitations & Future Work
